---
title: 'Reproducible Research: Peer Assessment 1'
author: "Tomaso Bulligan"
date: "Sunday, September 14, 2014"
output: html_document
---

```{r echo=FALSE}
library(knitr)
library(ggplot2)
opts_chunk$set(dev='CairoPNG')
```

# Reproducible Research: Peer Assessment 1

*Make sure to read READNE.md for additional information as to the scope of this research and instruction on how to obtain the data needed for the analysis.*

## Loading and preprocessing the data

We will load the data from the csv file provided with the assignment. Make sure said file has been extracted from its archive.

The date field will be converted to POSIX format for ease of analysis.

```{r}
activity <- read.csv("activity.csv", stringsAsFactors = FALSE)
activity$date <- as.Date(activity$date)
```

## What is mean total number of steps taken per day?

Next we will calculate and plot the total number of steps taken each day.

```{r}
# Calculate steps taken each day
daily.steps <- aggregate(steps ~ date, activity, sum, na.rm = TRUE)
names(daily.steps) <- c("date", "steps")

# Plot steps taken each day
qplot(data = daily.steps, x = date, y = steps, stat = "identity",
      geom = "bar") + labs(title = "Fig. 1: Number of steps per day",
                           x = "Date", y = "Steps")
```

Lastly we will compute the mean and the median of the total steps taken each day.

```{r}
mean.steps.day <- print(mean(daily.steps$steps, nsmall = 2))
median.steps.day <- print(median(daily.steps$steps))
```

## What is the average daily activity pattern?

With a methodology similar to that used to answer the previous question, we will calculate and plot the average steps taken per each 5-minute time interval.

```{r}
# Calculate average steps taken each interval
steps.intervals <- aggregate(steps ~ interval, activity, mean)
names(steps.intervals) <- c("interval", "steps")

# Plot average steps taken each interval
qplot(data = steps.intervals, x = interval, y = steps, geom = "line") +
    labs(title = "Fig. 2: Number of steps per time interval",
         x = "Interval", y = "Steps")
```

The interval with the maximum activity is the following:

```{r}
steps.intervals[which.max(steps.intervals$steps), ]
```

## Imputing missing values

First we will locate and quantify missing values.

```{r}
# Find what variables contain na and in what quantity
count.na <- apply(activity, 2, is.na)
apply(count.na, 2, sum)
```

We will fill missing values with the mean of the steps taken for every 5-minute interval.

```{r}
# Substitute missing values with mean of steps per interval
activity.filled <- merge(activity, steps.intervals, by = "interval")
activity.filled$steps <- ifelse(is.na(activity.filled$steps.x),
                                activity.filled$steps.y,
                                activity.filled$steps.x)
activity.filled <- activity.filled[ , c(1, 3, 5)]

# Plot steps taken each days with imputed missing values
qplot(data = activity.filled, x = date, y = steps, stat = "identity",
      geom = "bar") +
    labs(title = "Fig. 3: Number of steps per day, imputing missing values",
                           x = "Date", y = "Steps")
```

Lastly we will assess the impact of imputing missing values on the accuracy of our calculations.

```{r}
# Calculate steps taken each day with imputed missing values
daily.steps.filled <- aggregate(steps ~ date, activity.filled, sum, na.rm = T)

# Calculate mean and median of steps per day with imputed missing values
mean.steps.day.filled <- print(mean(daily.steps.filled$steps, nsmall = 2))
median.steps.day.filled <- print(median(daily.steps.filled$steps))

# Calculate percent difference between imputing and non imputing missing values
pct.diff <- function(x, y) {((x-y)/x)*100}
pct.diff(mean.steps.day.filled, mean.steps.day)
pct.diff(median.steps.day.filled, median.steps.day)
```

According to the above data, the impact of imputing missing values is negligible for our data set.

## Are there differences in activity patterns between weekdays and weekends?

The first step will be to identify which days fall on a weekday and which on a weekend. For this calculation the imputed values will be used.

```{r}
weekend.days <- c("Saturday", "Sunday")
activity.filled$daytype <- weekdays(activity.filled$date)
activity.filled$daytype <- ifelse(activity.filled$daytype %in% weekend.days,
                                  "weekend", "weekday")
activity.filled$daytype <- factor(activity.filled$daytype)
```

The following plot compares the steps taken in each 5-minute interval during weekdays and weekend days.

```{r}
# Calculate comparison of steps per interval between weekdays and weekend days
steps.intervals.filled <- with(activity.filled,
                              aggregate(steps, list(daytype, interval), mean))
names(steps.intervals.filled) <- c("daytype", "interval", "steps")

# Plot comparison of steps per interval between weekdays and weekend days
qplot(data = steps.intervals.filled, x = interval, y = steps, geom = "line") +
    labs(title = "Fig. 4: Number of steps per time interval per part of week",
                            x = "Interval", y = "Steps") +
    facet_grid(daytype ~ .)
```
